An Accurate Model for SARS-CoV-2 pooled RT-PCR Tests Errors

### 2021-03-16 comment.tex
New story, new structure

\documentclass{article}
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{todonotes}
\usepackage{amssymb}
\usepackage{subcaption}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{marginnote}
\usepackage[colorlinks=true,urlcolor=blue,citecolor=red,linkcolor=red,bookmarks=true]{hyperref}
%strikethrough
\usepackage{soul}
\usepackage{pgfplots}
\pgfplotsset{compat=1.14}
\usepackage{float}
\usepackage[margin=0.5in]{geometry}
\hypersetup{linkcolor=black,citecolor=black,urlcolor=black}

\newcommand{\fp}{P_{\text{fp}}}
\newcommand{\fn}{P_{\text{fn}}}
\newcommand{\Se}{S_e}
\newcommand{\Sp}{S_p}
\newcommand{\mi}{P_{\text{sfn}}}
\newcommand{\rel}{E_{\text{rel}}}
\renewcommand\Affilfont{\itshape\small}
\renewcommand{\Pr}{\mathbb{P}}

\begin{document}

\title{Inflated false-negative rates in pooled RT-PCR tests of SARS-CoV-2}

@@

\section{Introduction}

RT-PCR testing is a key component in breaking transmission chains and mitigating the COVID-19 pandemic. As such, the need for large-scale testing has resulted in the development of pooling schemes of RT-PCR tests \cite{DorfmanYuvalDor, PoolSize30, BayesianDorfman, MatrixPooling, LionDorfman}. One such popular scheme is Dorfman pooling \cite{DorfmanOriginal, DorfmanYuvalDor}: Select $N$ individuals and perform a single RT-PCR test on their combined (``pooled'') samples. If the pooled test yields a positive result --- test each individual separately. The throughput efficiency of Dorfman pooling has been demonstrated empirically \cite{DorfmanYuvalDor}. However, when test error rates are taken into consideration, a sharp increase in false-negative rates can be expected.

It is important to distinguish three types of false-negative events when performing pooling. For convenience, we follow a single \emph{infected} individual, henceforth referred to as "Donald". A \emph{single test's} false-negative is the event of a negative result upon testing Donald separately, i.e., in an RT-PCR test without pooling. We denote the test sensitivity $\Se$, so the probability of a single test false-negative is $1-\Se$. A \emph{pooled} false-negative occurs when a pooled test containing Donald's sample (and other samples) yields a negative result, i.e., the pooling fails to detect at least one positive result. Lastly, a \emph{scheme} false-negative occurs when an entire pooling scheme fails to identify Donald as infected. Our goal is to calculate Dorfman's scheme false-negative rate. Rephrasing, we wish to answer the following question: what is the probability of not identifying Donald as infected under a Dorfman pooling scheme?

\section{Methods}
\subsection{Probabilistic Assumptions}\label{subsec:assumptions}
We assume two pathways for a positive pooled test result: Viral RNA from an infected individual is correctly detected; or, some erroneous detection occurs (e.g. contaminant viral RNA is introduced). We ignore cross-reactivity with other Coronaviruses, which is negligible \cite{KitComparison}. We assume a homogeneous and disconnected population (each individual is infected independently and with equal probability). For simplicity, we do not take into account sample dilution, since it can only further increase false-negative rates \cite{DorfmanYuvalDor}.

Our assumptions, although natural, stand in contrast to assumptions commonly made in the literature \cite{Kim, Simplistic1, Simplistic2, OptimalDorfmanPool}. It is commonly assumed that the probability of a positive pooled test is not increased by having more than one infected individual in the pool. We refute the common assumptions with experimental data summarized in table \ref{table}, collected from \cite{Salazar}. There, the authors investigate Dorfman pooling and, regardless of the pooled test result, follow up and test each pool member separately. Here, we focus on 128 pools for which at least one subsequent separate test was positive --- of which 29 pooled tests were negative and 99 positive. In the data cited in \cite{Salazar}, of the 29 negative pools, subsequent separate testing yielded a single positive result in 24. In contrast, of the 99 positive pools, 42 yielded a single positive test upon subsequent separate testing.

@@

\subsection{calculation of Dorfman's scheme false-negative rate}\label{subsec:detailed}
Denote the prevalence of infection in the (tested) population $q$. As before, $\Se$ is the test's sensitivity, so $1-\Se$ is the single RT-PCR test's false-negative rate. We also denote $\Sp$ the test specificity --- the true-negative probability. This encompasses events of an erroneous RNA detection (via any possible pathway), which may cause a false-positive. By our assumptions, a pool containing Donald's sample and $N-1$ other samples will yield a \emph{negative} result if all of the following occur: 
\begin{itemize}
\item No erroneous detection occurred (i.e. no false-positive). This happens with probability $\Sp$.
\item The detection process fails for Donald's sample. A false-negative occurs for Donald, with probability $1-\Se$.
\item No detection for any of the other $N-1$ samples. For a single sample, the probability of being detected is the prevalence of SARS-CoV-2 in the tested population $q$, multiplied by the sensitivity $\Se$, hence the probability of detection is $q\Se$. For $N-1$ such samples, the probability of \emph{not} being identified is $(1-q\Se)^{N-1}$. 
\end{itemize}
The pooled false-negative probability for Donald is simply the product of the terms above. Hence:

\begin{align}
    \begin{split}
        \Pr(\text{pool is positive}) &= 1 - \Pr(\text{pool false-negative}) \\
        &= 1 - \Sp(1-\Se)(1-q\Se)^{N-1}.
    \end{split}
\end{align}

@@

\section{Results}\label{section:results}

>>>>

\documentclass{article}
\input{definitions}

\begin{document}
\title{An Accurate Model for SARS-CoV-2 pooled RT-PCR Tests Errors}

@@

\section{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in breaking transmission chains and mitigating the
COVID-19 pandemic. As such, the need for large-scale testing has
resulted in the development techniques that of pooling schemes of RT-PCR tests
\cite{DorfmanYuvalDor, PoolSize30, BayesianDorfman, MatrixPooling,
  LionDorfman}. One such popular scheme is Dorfman pooling
\cite{DorfmanOriginal, DorfmanYuvalDor}: Select $N$ individuals and
perform a single RT-PCR test on their combined (\emph{pooled})
samples. If the pooled test yields a positive result --- test each
individual separately. The throughput efficiency of Dorfman pooling
has been demonstrated empirically \cite{DorfmanYuvalDor}.

Pooling (or RT-PCR tests, for that matter) is not perfect, and there
is always some nonzero probability of error. Tyipically, authors
assume the probability of a true-positive (the test's
\emph{sensitivity}) does not depend on the number of infected samples
in the pool, but rather on the existence of at least one such
sample. Thus, the probability of a positive result in a pooled test is
(assumed) identical for a pool with one sample from an infected
individual and (e.g.) five such samples. This assumption is common in
the group testing literature \cite{Kim, OptimalDorfmanPool}, as well
as in more specific, COVID-19 focused studies \cite{Simplistic1,
  Simplistic2}. In this study we refute these commonly made
assumptions and show how they affect estimation of false-negative
rates for Dorfman pooling.

%% Before we proceed, it is important to distinguish three types of
%% false-negative events when performing pooling. For convenience, we
%% follow a single \emph{infected} individual, henceforth referred to as
%% "Donald". A \emph{single test's} false-negative is the event of a
%% negative result upon testing Donald separately, i.e., in an RT-PCR
%% test without pooling. We denote the test sensitivity $\Se$, so the
%% probability of a single test false-negative is $1-\Se$. A
%% \emph{pooled} false-negative occurs when a pooled test containing
%% Donald's sample (and other samples) yields a negative result, i.e.,
%% the pooling fails to detect at least one positive result. Lastly, a
%% \emph{scheme} false-negative occurs when an entire pooling scheme
%% fails to identify Donald as infected. Our goal is to calculate
%% Dorfman's scheme false-negative rate. Rephrasing, we wish to answer
%% the following question: what is the probability of not identifying
%% Donald as infected under a Dorfman pooling scheme?

\subsection{RT-PCR}
Before delving into our probabilistic assumptions, we briefly explain
the process of testing for SARS-CoV-2 by RT-PCR. Our exposition
intentionally avoids many details and a comprehensive review of RT-PCR
can be found in \cite{COVID-RTPCR, PCRBook}.

PCR is a process in which a targeted DNA strand's frequency is
amplified to create billions of copies in a reaction mixture. At the
end of this \emph{amplification} process, the presence of targeted DNA
molecules in a reaction mixture can be confidently
determined. Arguably, the best way to understand PCR is as a chain
reaction: Each targeted DNA molecule is copied into two, that are
copied into four and so forth. This cascade gives PCR its name:
polymerase chain reaction.

Specifically, the PCR process is comprised of cycles, with each cycle
doubling the abundance of the targeted DNA. In each cycle, each
double-strand DNA molecule is broken into two separate strands. An
enzyme called DNA polymerase generates a new double-strand DNA replica
from each separate strand. Replication cannot start without a specific
short DNA sequence, called a primer, that has to be introduced into
the reaction mixture. Introducing the right primer into the reaction
mixture ensures (almost) only the targeted DNA sequence is copied. A
protein is added to the reaction mixture, which emits light upon
successfully binding the targeted DNA. Once enough light is emitted
the tested reaction mixture is declared positive and a
\emph{detection} is said to occur. If, on the other hand, a predefined
number of cycles pass without a detection event then the reaction
mixture is declared negative.

Since the genetic material of SARS-CoV-2 is RNA rather than DNA, an
extra preprocessing step is required. An enzyme called
reverse transcription (hence the RT in RT-PCR) replicates existing RNA
in the reaction mixture into DNA. Once DNA replicas are made, the PCR
process proceeds as described above, targeting SARS-CoV-2 DNA
replicas.

\section{RT-PCR errors}
Failed detection of SARS-CoV-2 RNA in pooled RT-PCR testing is
referred to below as a false-negative. One possible source of
false-negatives in pooling is sample dilution. When pooling, several
samples are mixed, so the concentration of viral RNA is reduced. This
effect may cause a delay in amplification \cite{DorfmanYuvalDor}, no
detection and, consequently, a false-negative. However, \cite{Lion}
showed that this effect can be safely ignored when mixing up to 32
samples, which we correspondingly assume.

\subsection{Common assumptions}
Previous studies of group testing strategies assumed that the
false-negative probability does not depend on the number of infected
samples, but merely on the existence of at least one such sample in a
pool \cite{Kim, OptimalDorfmanPool}. Current studies of pooling in the
context of SARS-CoV-2 also employ similar assumptions
\cite{Simplistic1, Simplistic2}. Specifically, these studies assume
that the probability of a negative result is the same for a pool with
a single sample coming from an infected individual and (e.g.)
five. We refute these assumptions and suggest more plausible ones in
Section~\ref{subsec:ours} below.

\subsection{Our model}\label{subsec:ours}
We assume two pathways for a positive pooled test result: Viral RNA
from an infected individual is correctly detected; or, some erroneous
detection occurs (e.g. contaminant viral RNA is introduced). We assume
a homogeneous and disconnected population (each individual is infected
independently and with equal probability).

Our assumptions, although natural, stand in contrast to assumptions
commonly made in the literature \cite{Kim, Simplistic1, Simplistic2,
  OptimalDorfmanPool}. It is commonly assumed that the probability of
a positive pooled test is not increased by having more than one
infected individual in the pool. We refute the common assumptions with
experimental data summarized in Table \ref{table}, collected from
\cite{Salazar}. There, the authors investigate Dorfman pooling and,
regardless of the pooled test result, follow up and test each pool
member separately. Here, we focus on 128 pools for which at least one
subsequent separate test was positive --- of which 29 pooled tests
were negative and 99 positive. In the data cited in \cite{Salazar}, of
the 29 negative pools, subsequent separate testing yielded a single
positive result in 24. In contrast, of the 99 positive pools, 42
yielded a single positive test upon subsequent separate testing.

@@

Erroneous detection of SARS-CoV-2 RNA (a false-positive) in pooled
RT-PCR testing can also occur. A common assumption \cite{Simplistic1,
  Simplistic2, Kim, OptimalDorfmanPool} is that the false-positive
probability does not depend on the number of negative samples in a
pool. We incorporate this assumption in our likelihood model with a
small modification. We assume that an erroneous amplification can
occur in \emph{any} pool. Specifically, it is possible that correct
amplifications fail and an erroneous one occurs simultaneously. This
assumption is relatively specific for the current application of
screening for SARS-CoV-2 via RT-PCR.  For example, cross-reactivity
with other coronaviruses would have violated this assumption, but it
was ruled out in \cite{KitComparison}.

\begin{figure}[H]
  \centering
  \begin{tikzpicture}
    \begin{scope}[every node/.style={circle,thick,draw}]
      \node[shape=circle,draw=black] (A) at (0,0) {$\theta_1=1$};
      \node[shape=circle,draw=black] (B) at (2,0) {$\theta_2=0$};
      \node[shape=circle,draw=black] (D) at (4,0) {$\theta_3=0$};
      \node[shape=circle,draw=black] (E) at (6,0) {$\theta_4=1$};

      \node[shape=ellipse,draw=black]
      (F) at (9,0) {False-positive};

      \node[shape=rectangle,draw=black] (G) at (4,7)
           {$\Pr(\data=0|\theta)=\fn^2(1-\fp)$};
    \end{scope}

    \begin{scope}[>={Stealth[black]},
        every node/.style={fill=white},
        every edge/.style={draw=red,very thick}]
      \path [->] (A) edge node {$\fn$} (4.5,6.6);
      %\path [->] (B) edge node {$0$} (G);
      %\path [->] (D) edge node {$0$} (G);
      \path [->] (E) edge node {$\fn$} (4.5,6.6);
      \path [->] (F) edge node {$1-\fp$} (5.5,6.6); 
    \end{scope}
  \end{tikzpicture}
  \caption{Illustration of the likelihood model of a pooled test
    result. A pool contains individuals $\{1,2,3,4\}$ with state
    $\theta=(1,0,0,1)$ (i.e. individuals 1 and 4 are infected). A
    negative pooled test implies that three detection paths failed. A
    false negative occurred for $1$ and $4$, each with probability
    $\fn$. Additionally, no erroneous detection of SARS-CoV-2
    occurred, with probability $1-\fp$. Individuals $2$ and $3$ are
    not infected and do not contribute to the probability of the
    pooled test result. A positive pooled test arises if any one of
    the above mentioned paths results in a
    detection.}\label{fig:likelihood}
\end{figure}

To summarize, we assume that for a single pool, a positive test result
is generated in one of two paths. Either, SARS-CoV-2 RNA from an
infected individual's sample is correctly amplified and detected, and
this can happen for each positive sample in a pool. Or, some erroneous
amplification occurs (e.g. contaminant viral RNA is introduced), an
event that occurs at most once per pool. Our model is illustrated in
Figure~\ref{fig:likelihood}, and explicitly defined below.

Consider a pool containing $N$ individuals $\{1,\dots,N\}$. We denote
the true infection states $\theta \in \{0,1\}^N$, so individual $i$ is
infected iff $\theta_i=1$. Data are denoted $\data$. The tests
false-negative rate is denoted $\fn$ and the false-positve rate is
denoted $\fp$. The probability of a negative pooled test result is
presented in \eqref{eq:likelihood a} (along with its complement
\eqref{eq:likelihood b}), and explained below.
\begin{subequations}
\begin{align}%\label{eq:single likelihood}
  %\begin{split}
    \Pr(\data=0 | \theta) &= (1-\fp)\prod_{i=1}^N
    \fn^{\theta_i}, \label{eq:likelihood a}\\
    %
    %
    %
    \Pr(\data=1 | \theta) &= 1 - (1-\fp)\prod_{i=1}^N \fn^{\theta_i}. \label{eq:likelihood b}
  %\end{split}
\end{align}
\end{subequations}
A negative pooled test result occurs when all detection paths (both
correct and erroneous) fail. The probability of no false-detection
accounts for the $1-\fp$ term. The probability of no correct detection
is $\fn$ per infected individual. The probability that all such paths
fail is the product of the above mentioned terms, displayed in
\eqref{eq:likelihood a}. The probability of a positive result,
presented in \eqref{eq:likelihood b}, is simply the complement.
Combining \eqref{eq:likelihood a} and \eqref{eq:likelihood b}, and
recalling that $\data \in \{0,1\}$ yields:

\begin{align}\label{eq:combined likelihood}
  \begin{split}
    \Pr(\data| \theta) &= \left [(1-\fp)\prod_{i=1}^N
      \fn^{\theta_i} \right ]^{1-\data} \left [1 - (1-\fp)\prod_{i=1}^N \fn^{\theta_i} \right ]^{\data}.
  \end{split}
\end{align}

\section{Application: Calculation of Dorfman's scheme false-negative rate}\label{subsec:detailed}
As an application of our model, we present the differnece in scheme
false-negative rates (...) Denote the prevalence of infection in the
(tested) population $q$. As before, $\Se$ is the test's sensitivity,
so $1-\Se$ is the single RT-PCR test's false-negative rate. We also
denote $\Sp$ the test specificity --- the true-negative
probability. This encompasses events of an erroneous RNA detection
(via any possible pathway), which may cause a false-positive. By our
assumptions, a pool containing Donald's sample and $N-1$ other samples
will yield a \emph{negative} result if all of the following occur:
\begin{itemize}
\item No erroneous detection occurred (i.e. no false-positive). This
  happens with probability $\Sp$.
\item The detection process fails for Donald's sample. A
  false-negative occurs for Donald, with probability $1-\Se$.
\item No detection for any of the other $N-1$ samples. For a single
  sample, the probability of being detected is the prevalence of
  SARS-CoV-2 in the tested population $q$, multiplied by the
  sensitivity $\Se$, hence the probability of detection is $q\Se$. For
  $N-1$ such samples, the probability of \emph{not} being identified
  is $(1-q\Se)^{N-1}$.
\end{itemize}
The pooled false-negative probability for Donald is simply the product of the terms above. Hence:

\begin{align}
    \begin{split}
        \Pr(\data=1|\theta_1=1) &= 1 - \Pr(\data=0|\theta_1=1) \\
        &= 1 - \Sp(1-\Se)(1-q\Se)^{N-1}.
    \end{split}
\end{align}

@@

\subsection{Results}\label{section:results}
